### 2014-10-21 paper/include/flikker.tex
corrections

%flikker
The technique is proposed by Liu et al. \cite{flikker} and targets Data Tolerance and Validity. Flikker distinguish data as critical or non-critical. The critical data is refreshed at the normal maximal refresh period, whereas the non-critical data is refreshed at a much lower rate. Refreshing at a lower rate saves refreshing power but increases data corruption. This trade-off was found to be balanced with a refresh period of one second, where the refresh power savings was expected to be $23\%$ and the error rate $4.0 \times 10^{-8}$, if the critical part is $25\%$.

Flikker refreshes the data by extending the \textit{Partial Array Self Refresh} (PASR) mode which was introduced 2002 in mobile DRAMs. PASR mode refreshes a portion of the DRAM array of one bank when the mobile system is in sleep mode, and the technique modifies PASR's internal address counter to support two refresh rates.

>>>>

%flikker
The technique is proposed by Liu et al. \cite{flikker} and targets Data Tolerance. Flikker distinguish data as critical or non-critical. The critical data is refreshed at the normal maximal refresh period, whereas the non-critical data is refreshed at a much lower rate. Refreshing at a lower rate saves refreshing power but increases data corruption. This trade-off was found to be balanced with a refresh period of one second, where the refresh power savings was expected to be $23\%$ and the error rate $4.0 \times 10^{-8}$, if the critical part is $25\%$.

Flikker refreshes the data by extending the \textit{Partial Array Self Refresh} (PASR) mode which was introduced 2002 in mobile DRAMs. PASR mode refreshes a portion of the DRAM array of one bank when the mobile system is in sleep mode, and the technique modifies PASR's address counter to support two refresh rates.

### 2014-10-21 paper/include/raidr.tex
corrections

In this approach, RAIDR by Liu et al. \cite{raidr}, all memory rows gets divided into different groups, called bins, based on the cell with the shortest Retention Time in the row. In the basic configuration there are two bins, one with rows that need a refresh rate of $64\:ms \to 128\:ms$ and one with rows that need to be refreshed every $128\:ms \to 256\:ms$. Rows in the first bin gets refreshed every $64\:ms$, rows in the second bin every $128\:ms$, and all remaining rows gets refreshed at the default refresh rate which is set to $256\:ms$. This is done using two counters, one row counter that rolls over every $64\:ms$, and one counter that keeps track on how many times the row counter have rolled over. With this every row will get selected as refresh candidate every $64\:ms$, the refresh will be performed depending on which bin the row belongs to and on the second counter's value. The actual refreshes are performed by activating the row in question, essentially performing a ROR. RAIDR conform to temperature variations by utilizing a third counter that is used to scale all bins refresh rate with the temperature.

The implementation is done inside the memory controller meaning RAIDR can not be adopted in currently existing systems. To minimize the storage needed for keeping track of bin members Bloom filters is used. Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration. The storage requirement increases together with the number of bins used while the granularity, and thereby the decrease in refresh operations, of the approach becomes higher with more bins. As well as the number of bins, the size chosen for the Bloom filters also affects the storage overhead. Lie et al. evaluates twelve different configurations ranging from one bin with small Bloom filters to three bins with large Bloom filters. The later gives the highest decrease in refresh operations but needs \textit{65.25 KB} memory while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed. The default configuration is a trade of between these two extremes.

>>>>

In this approach, RAIDR by Liu et al. \cite{raidr}, all memory rows get divided into different groups, called bins, based on the cell with the shortest Retention Time in the row. In the basic configuration there are two bins, one with rows that need a refresh rate of $64\:ms \to 128\:ms$ and one with rows that need to be refreshed every $128\:ms \to 256\:ms$. Rows in the first bin gets refreshed every $64\:ms$, rows in the second bin every $128\:ms$, and all remaining rows gets refreshed at the default refresh rate which is set to $256\:ms$. This is done using two counters, one address counter that rolls over every $64\:ms$, and one counter that keeps track on how many times the address counter have rolled over. With this, every row will get selected as refresh candidate every $64\:ms$ and the refresh will be performed depending on which bin the row belongs to and on the second counter's value. The actual refreshes are performed by activating the row in question, essentially performing a ROR. RAIDR conform to temperature variations by utilizing a third counter that is used to scale all bins refresh rate with the temperature.

The implementation is realized by modifying the MC. To minimize the storage needed for keeping track of bin members, Bloom filters are used. Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration, which becomes 0.037 ppm. The storage requirement increases together with the number of bins used, while the granularity, thereby the decrease in refresh operations, of the technique becomes higher with more bins. Similarly to the number of bins, the size chosen for the Bloom filters also affects the storage overhead.

Liu et al. evaluates twelve different configurations ranging from one bin with small Bloom filters to three bins with large Bloom filters. The latter gives the highest decrease in refresh operations but needs \textit{65.25 KB} of memory, while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed. The default configuration is a trade of between these two extremes.

### 2014-10-21 paper/include/sparkk.tex
corrections

The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by the internal address counter. Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank. An entry consists of the chosen refresh period for each bank, time-out bits for each bank, and the address to the end of the memory area. When the counter address and the address to the end of the memory area matches, the MC advances to the next entry. There, the MC decides if to issue a refresh operation and if so, which banks to send the command to. The DRAM device has to provide the ability of per bank refresh commands for this to work.

\todo[inline]{Drafty below}

Similar to Flikker, Sparkk needs software support to partition the data. As Sparkk is a more fine grained technique then Flikker, a more extensive support is needed. Lucas et al. proposes a complex .....

Results:

>>>>

The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by a address counter. Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank. An entry consists of the chosen refresh period for each bank, time-out bits for each bank, and the address to the end of the memory area. When the counter address and the address to the end of the memory area matches, the MC advances to the next entry. There, the MC decides if to issue a refresh operation and if so, which banks to send the command to. The DRAM device has to provide the ability of per bank refresh commands for this to work. By this scheme, Sparkk only refreshes valid data and thus utilize Validity information.

Similar to Flikker, Sparkk needs software support to partition the data. As Sparkk is a more fine grained technique then Flikker, a more extensive support is needed. Lucas et al. present a vision where the technique is part of a tailored system for approximate data with specialized languages, instruction set, and hardware.

Through stochastic modeling, Sparkk has over $10~dB$ higher peak SNR than Flikker at the same average refresh rate. Moreover, Sparkk produces higher peak SNR at half the average refresh rate used for Flikker. Experiments regarding energy savings has unfortunately not been performed.
